---
title: "BLUP vs Linear HFA"
author: "PME"
date: "9/16/2021"
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding=encoding, output_dir=here::here('results'))})
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: paged
    code_folding: hide
  pdf_document: default
  word_document: default
editor_options: 
  chunk_output_type: inline
---

# Overview
Comparison of BLUP (MacQueen et al, in review) and linear (Ewing et al, 2019) ways
of identifying home sites and thereby calculating the home field advantage.

Comparison will:
a) compare HFA for each entry
b) recreate Figure 2b from MacQueen et al (in review).

## Setup
### Paths
```{r}
in_dir = 'data'  # within the R project
in_perf = 'Phenotypes_of_sequenced_individuals_home_away_analysis.csv'  # yield data

out_dir = 'results'
```



### Variables
```{r}
# ids of columns, etc
site = 'Location_code'  # spatial environment
year = 'Year'           # temporal environment
geno = 'Taxa'           # genotype
pheno = 'SY'            # phenotype of interest = yield. Units kg/ha.
popn = 'Race'           # population
race = c('Durango', 'Mesoamerican', 'Nueva Granada')
names(race) = race

seed = 4456  # for permutations
permute_times = 999

color_palette='FantasticFox1'

# nice names for model terms
prettynames = c(
  is_home = 'Home Site',
  Residuals = 'Residuals',
  Taxa = 'Genotype',
  Location_code = 'Site',
  Year = 'Year',
  `Year:Location_code` = 'Site-Year',
  GxE.Reduction = 'GxE.Reduction'
)
```

#### Export?
Export results to figures, rdata, etc?
```{r load}
export = FALSE
export = TRUE
```

### Libraries and functions
```{r libraries}
libs = c(
## workflow and scripting
  'here',      
  'magrittr', 
  'reshape2', 
  'broom',
  'parallel',
## modeling
  'lme4',     
  'car', 
  'agricolae',
  'vegan', 
## Plotting
  'wesanderson',
  'ggplot2', 
  'ggpmisc',
  'gridExtra',
  'ggtext',
  'glue',
## tables
  'gt'
)

fn_path = file.path(getwd(), 'R', 'functions')
for (i in list.files(fn_path, full.names=TRUE)) {
  source(i)
}

for (i in libs) get_libraries(i)  # install if necessary
```

### Load and munge

Main dataframe
```{r}
df = here(in_dir, in_perf) %>% 
  read.csv(stringsAsFactors=FALSE)

# Merge the Jalisco race into Durango (which is supported by genetics)
df[, popn] %<>% gsub('Jalisco', 'Durango', .)

# Rename the Nueva Granada race to Andean
# df[, popn] %<>% gsub('Nueva Granada', 'Andean', .)

# make a numeric version of year for temporal trends.
year_num = paste(year, 'Num', sep='_')
df[, year_num] = df[, year] - max(df[, year])

# year is otherwise a factor
df[, year] %<>% 
  as.numeric %>% 
  factor(., levels=sort(unique(.)))

# filter siteyears
min_siteyears = 3
df %<>% filter_siteyears(site='Location_code',
                         year='Year',
                         min_times=min_siteyears)

type.convert(df) %>% summary
```

Colors
```{r}
colors = wes_palette(color_palette)
colors
```


### Notebook Functions
```{r}
# Pretty print values
pp = function(x, digits=3, decimals=2) {
  sprinter = paste('%.', decimals, 'f', sep='')
  value = signif(x, digits)
  out = sprintf(sprinter, value)
  return(out)
}

# Standard Error
se = function(x) sd(x) / sqrt(length(x))

# tidy a linear model
statsum = function(x, coef, stat=c('est', 'se', 'tstat', 'pval')) {
  out = x[x$term==coef, ]
  stat = switch(stat,
                'est' = out[, 'estimate'],
                'se' = out[, 'std.error'],
                'tstat' = out[, 'statistic'],
                'pval' = out[, 'p.value'])
  return(out)
}

# munge an anova table
get_ss = function(model) {
  a = model
  out = data.frame(
    Predictor = rownames(a),
    `Sum Squared` = a$`Sum Sq`,
    `Proportion of Variance` = round(a$`Sum Sq` / sum(a$`Sum Sq`) *100, 2),
    df = a$Df,
    `F Statistic` = round(a$`F value`, 4),
    `p Value` = signif(a$`Pr(>F)`, 3)
  )
  return(out)
}
gg_themer = function(x, remove_strip=FALSE) {
  
  out = x + 
  theme_minimal() +
  theme(panel.grid=element_blank(),
        axis.ticks=element_line(color='gray',
                                size=0.2),
        axis.title=element_text(size=9))
  
  if (remove_strip) {  # for the bottom panel of fully faceted figures
    out = out + theme(strip.text=element_blank())
  }
  return(out)
}

gg_mapthemer = function(x) {
  out = x + 
    theme_minimal() +
    theme(axis.ticks=element_line(color='gray',
                              size=0.2),
      axis.title=element_text(size=9),
      legend.title=element_text(size=9),
      panel.grid=element_line(color='gray',
                              size=0.2),
      legend.position='bottom')
  return(out)
}
```


# BLUP
Uses a mixed effects model to shrink average performance at each site for each variety toward the overal site mean. Rare sites are shrunk more agressively. This reduces noise and the influence of rare sites, balancing raw performance with our confidence in the estimate.
```{r}
blup_df = df
blup_df$is_home = NULL
pop_hfa_blup = permute_hfa(blup_df, level='population', 
                           site, year, geno, pheno, popn, 
                           times=permute_times, seed=seed, blup_home=TRUE)
```


```{r}
cbind(observed = pop_hfa_blup$perms[, 1], 
      expected = pop_hfa_blup$perms[, 1] - pop_hfa_blup$home_field[, 'median'],
      pop_hfa_blup$home_field)
```
Andean HFA was not larger than expected by chance. Durango and Mesoamerican lines both received more HFA benefit than expected. 

Does this change across time?

```{r}
temp_hfa_blup = temporal_hfa(blup_df, site, year, geno, pheno, popn, blup_home=TRUE)
```

### TABLE S3: HFA ~ Time Anova
```{r}
temp_hfa_blup$anova
```
Year matters for some races

```{r}
tabs3_data = temp_hfa_blup$anova
rownames(tabs3_data) %<>%
  gsub('_num', '', .) %>% 
  gsub('\\:', '*', .)
tabs3 = tabs3_data %<>%
  data.frame(X = rownames(.),
             .) %>% 
  set_colnames(c(' ', colnames(tabs3_data))) %>% 
  gt %>% 
  fmt_number(columns=c('Sum Sq'), n_sigfig=4) %>% 
  fmt_number(columns='F value', decimals=2) %>% 
  fmt_number(columns='Pr(>F)', decimals=3) %>% 
  fmt_missing(columns=c(4, 5), missing_text='-') %>% 
  tab_header('Table S3 (BLUP HFA)',
             subtitle='ANOVA of Home Field Advantage Across Time') %>% 
  opt_align_table_header(align='left') %>% 
  cols_align('right',
             columns=' ')
  
tabs3
```

```{r}
temp_hfa_blup$model %>% summary
```

## Plot
First generate the supplementary datasets
```{r}
thfa_df = temp_hfa_blup$temporal_hfa
# format x axis
year_num = paste0(year, '_num')
year_num0 = gsub('num', 'num0', year_num)

thfa_df[, year_num0] = thfa_df[, year_num] - min(thfa_df[, year_num])

xstep = 10
xrange = thfa_df[, year_num] %>% 
  range
xmin = xrange[1] %>% 
  divide_by(xstep) %>% 
  floor %>% 
  multiply_by(xstep)
xmax = xrange[2]
xbreaks = seq(xmin, xmax, by=xstep)
xbreaks_minor = xbreaks + (xstep/2)

# generate formula annotations for each model
ff = temp_hfa_blup$formulas$anova %>% 
  gsub('\\*', '', .) %>% 
  gsub(popn, '', .) %>%
  gsub(year_num, year_num0, .) %>% 
  formula
annotate = split(thfa_df, thfa_df[, popn]) %>% 
  sapply(function(x) {
    mod = lm(ff, data=x) %>% 
      summary %>% 
      coef
    p_val = mod[2, 4]
    # if (p_val <= 0.05) {
      f1 = paste('italic(y)==', 
                 signif(mod[1, 1], 3), '~+~', 
                 signif(mod[2, 1], 3), '~italic(x)', 
                 sep='') 
      f2 = ifelse(p_val < 0.001,
                  'italic(p)~<~0.001',
                  paste('italic(p)==', signif(p_val, 1)))
      out = paste('list(', f1, ',', f2, ')', sep='')
    # } else {
      # out = 'n.s.' 
    # }
    return(out)
  })
annotate =
  data.frame(popn = names(annotate),
             text = annotate,
             x = 1980,
             y = 1400)
names(annotate) %<>% 
  gsub('popn', popn, .)

# generate ribbons for expected values
exp_hfa = pop_hfa_blup$perms %>% 
  .[, -c(1)] %>% 
  apply(1, function(x) {
    c(p50 = median(x),
      p05 = quantile(x, 0.05) %>% set_names(NULL),
      p95 = quantile(x, 0.95) %>% set_names(NULL))
  }) %>% 
  t
exp_hfa %<>% data.frame(
  popn = rownames(.),
  .
) %>% 
  set_rownames(NULL) %>% 
  rbind(., .)
exp_hfa$x_val = c(
  rep(xrange[1], nrow(exp_hfa)/2),
  rep(xrange[2], nrow(exp_hfa)/2)
)
names(exp_hfa) %<>% gsub('popn', popn, .)
```


```{r fig.height=3, fig.width=6.5}
thfa_df %<>% .[, c(paste0(year, '_num'), 
                   paste0(year, '_num0'),
                   'Estimate', popn, 'Std.Error')]
blup_plot = thfa_df %>% 
  ggplot(aes_string(x=paste0(year, '_num'),
                    y='Estimate',
                    color=popn,
                    fill=popn,
                    grp.label=popn)) +
  ylim(-900, 2000) +
  facet_wrap(paste('~', popn) %>% formula,
             nrow=1) +
  scale_color_manual(values=colors, guide=NULL) +
  scale_fill_manual(values=colors, guide=NULL) +
  scale_x_continuous(breaks=xbreaks, minor_breaks=xbreaks_minor) +
  geom_hline(aes(yintercept=0),
             size=0.2,
             color='gray') +
  geom_ribbon(data=exp_hfa,
              aes_string(ymin='p05',
                         ymax='p95',
                         x='x_val'),
              fill='#EEEEEE',
              alpha=1,
              inherit.aes=FALSE) +
  geom_pointrange(aes(ymin=Estimate - Std.Error,
                      ymax=Estimate + Std.Error),
                  shape=20,
                  size=0.2) +
  geom_smooth(method='lm',
              formula=y~x,
              alpha=0.1,
              size=0.2,
              level=0.95,
              fullrange=TRUE) +
  geom_text(data=annotate,
            aes(x=x,
                y=y,
                label=text),
            color='black',
            hjust='inward',
            vjust='inward',
            parse=TRUE,
            size=3) +
  labs(x='Year',
       y='Home Field Advantage [kg/ha]',
       title='BLUP Home',
       color=NULL) +
  theme_minimal() +
  theme(#panel.grid=element_blank(),
        axis.title=element_text(size=9),
        axis.ticks=element_line(color='gray',
                                size=0.2))
blup_plot
```

# Linear
Uses the raw average performance at each site for each variety to ID home.
```{r}
lin_df = df
lin_df$is_home = NULL
pop_hfa_lin = permute_hfa(lin_df, level='population', 
                           site, year, geno, pheno, popn, 
                           times=permute_times, seed=seed, blup_home=FALSE)
```


```{r}
cbind(observed = pop_hfa_lin$perms[, 1], 
      expected = pop_hfa_lin$perms[, 1] - pop_hfa_lin$home_field[, 'median'],
      pop_hfa_lin$home_field)
```
Andean HFA was not larger than expected by chance. Durango and Mesoamerican lines both received more HFA benefit than expected. 

Does this change across time?

```{r}
temp_hfa_lin = temporal_hfa(lin_df, site, year, geno, pheno, popn, blup_home=FALSE)
```

### TABLE S3: HFA ~ Time Anova
```{r}
temp_hfa_lin$anova
```
Year matters for some races

```{r}
tabs3_data = temp_hfa_lin$anova
rownames(tabs3_data) %<>%
  gsub('_num', '', .) %>% 
  gsub('\\:', '*', .)
tabs3 = tabs3_data %<>%
  data.frame(X = rownames(.),
             .) %>% 
  set_colnames(c(' ', colnames(tabs3_data))) %>% 
  gt %>% 
  fmt_number(columns=c('Sum Sq'), n_sigfig=4) %>% 
  fmt_number(columns='F value', decimals=2) %>% 
  fmt_number(columns='Pr(>F)', decimals=3) %>% 
  fmt_missing(columns=c(4, 5), missing_text='-') %>% 
  tab_header('Table S3 (lin HFA)',
             subtitle='ANOVA of Home Field Advantage Across Time') %>% 
  opt_align_table_header(align='left') %>% 
  cols_align('right',
             columns=' ')
  
tabs3
```

```{r}
temp_hfa_lin$model %>% summary
```

## Plot
First generate the supplementary datasets
```{r}
thfa_df = temp_hfa_lin$temporal_hfa
# format x axis
year_num = paste0(year, '_num')
year_num0 = gsub('num', 'num0', year_num)

thfa_df[, year_num0] = thfa_df[, year_num] - min(thfa_df[, year_num])

xstep = 10
xrange = thfa_df[, year_num] %>% 
  range
xmin = xrange[1] %>% 
  divide_by(xstep) %>% 
  floor %>% 
  multiply_by(xstep)
xmax = xrange[2]
xbreaks = seq(xmin, xmax, by=xstep)
xbreaks_minor = xbreaks + (xstep/2)

# generate formula annotations for each model
ff = temp_hfa_lin$formulas$anova %>% 
  gsub('\\*', '', .) %>% 
  gsub(popn, '', .) %>%
  gsub(year_num, year_num0, .) %>% 
  formula
annotate = split(thfa_df, thfa_df[, popn]) %>% 
  sapply(function(x) {
    mod = lm(ff, data=x) %>% 
      summary %>% 
      coef
    p_val = mod[2, 4]
    # if (p_val <= 0.05) {
      f1 = paste('italic(y)==', 
                 signif(mod[1, 1], 3), '~+~', 
                 signif(mod[2, 1], 3), '~italic(x)', 
                 sep='') 
      f2 = ifelse(p_val < 0.001,
                  'italic(p)~<~0.001',
                  paste('italic(p)==', signif(p_val, 1)))
      out = paste('list(', f1, ',', f2, ')', sep='')
    # } else {
      # out = 'n.s.' 
    # }
    return(out)
  })
annotate =
  data.frame(popn = names(annotate),
             text = annotate,
             x = 1980,
             y = 1400)
names(annotate) %<>% 
  gsub('popn', popn, .)

# generate ribbons for expected values
exp_hfa = pop_hfa_lin$perms %>% 
  .[, -c(1)] %>% 
  apply(1, function(x) {
    c(p50 = median(x),
      p05 = quantile(x, 0.05) %>% set_names(NULL),
      p95 = quantile(x, 0.95) %>% set_names(NULL))
  }) %>% 
  t
exp_hfa %<>% data.frame(
  popn = rownames(.),
  .
) %>% 
  set_rownames(NULL) %>% 
  rbind(., .)
exp_hfa$x_val = c(
  rep(xrange[1], nrow(exp_hfa)/2),
  rep(xrange[2], nrow(exp_hfa)/2)
)
names(exp_hfa) %<>% gsub('popn', popn, .)
```


```{r fig.height=3, fig.width=6.5}
thfa_df %<>% .[, c(paste0(year, '_num'), 
                   paste0(year, '_num0'),
                   'Estimate', popn, 'Std.Error')]
lin_plot = thfa_df %>% 
  ggplot(aes_string(x=paste0(year, '_num'),
                    y='Estimate',
                    color=popn,
                    fill=popn,
                    grp.label=popn)) +
  facet_wrap(paste('~', popn) %>% formula,
             nrow=1) +
  ylim(-900, 2000) +
  scale_color_manual(values=colors, guide=NULL) +
  scale_fill_manual(values=colors, guide=NULL) +
  scale_x_continuous(breaks=xbreaks, minor_breaks=xbreaks_minor) +
  geom_hline(aes(yintercept=0),
             size=0.2,
             color='gray') +
  geom_ribbon(data=exp_hfa,
              aes_string(ymin='p05',
                         ymax='p95',
                         x='x_val'),
              fill='#EEEEEE',
              alpha=1,
              inherit.aes=FALSE) +
  geom_pointrange(aes(ymin=Estimate - Std.Error,
                      ymax=Estimate + Std.Error),
                  shape=20,
                  size=0.2) +
  geom_smooth(method='lm',
              formula=y~x,
              alpha=0.1,
              size=0.2,
              level=0.95,
              fullrange=TRUE) +
  geom_text(data=annotate,
            aes(x=x,
                y=y,
                label=text),
            color='black',
            hjust='inward',
            vjust='inward',
            parse=TRUE,
            size=3) +
  labs(x='Year',
       y='Home Field Advantage [kg/ha]',
       title='Raw Average Home',
       color=NULL) +
  theme_minimal() +
  theme(#panel.grid=element_blank(),
        axis.title=element_text(size=9),
        axis.ticks=element_line(color='gray',
                                size=0.2))
lin_plot
```

# Comparison at Entry Level
```{r}
process_permute = function(x, obs_name=c('BLUP', 'Average')) {
  obs_name = match.arg(obs_name)
  df = x$home_field[, c('Race', 'Taxa')]
  obs = x$perms %>% 
    do.call(rbind, .)
  df[, obs_name] = obs[df[, 'Taxa'], 1]
  return(df)
}

shrink_hfa = permute_hfa(df, 
                         level='genotype', 
                         site, 
                         year, 
                         geno, 
                         pheno, 
                         popn, 
                         times=1, 
                         blup_home=TRUE) %>% 
  process_permute(obs_name='BLUP')

lin_hfa = permute_hfa(df, 
                      level='genotype', 
                      site, 
                      year, 
                      geno, 
                      pheno, 
                      popn, 
                      times=1, 
                      blup_home=FALSE) %>% 
  process_permute(obs_name='Average')
```

```{r}
geno_hfa_compare = merge(shrink_hfa, lin_hfa, by=c(popn, geno))

geno_hfa_compare_plt = ggplot(geno_hfa_compare,
                              aes(x=Average,
                                  y=BLUP,
                                  color=Race)) +
  scale_color_manual(values=colors) +
  coord_equal() +
  geom_hline(aes(yintercept=0),
             size=0.2,
             color='darkgray') +
  geom_vline(aes(xintercept=0),
             size=0.2,
             color='darkgray') +
  geom_point(size=0.5) +
  geom_smooth(aes(x=Average,
                  y=BLUP), 
              inherit.aes=FALSE,
              method='lm',
              formula=y~x,
              size=0.2,
              level=0.95,
              fullrange=TRUE,
              alpha=0.1,
              color='darkgray') +
  geom_abline(aes(intercept=0,
                  slope=1),
              linetype='dashed',
              size=0.2,
              color='darkgray') +
  labs(x='Raw Average Home',
       y='BLUP Home',
       main='HFA of Each Entry [kg/ha]') +
  theme_minimal() %>% 
  gg_themer
geno_hfa_compare_plt +
  theme(legend.title=element_text(size=8))
```

# Combine and export
```{r fig.height=9, fig.width=6.5}
p1 = blup_plot
p2 = lin_plot
p3 = geno_hfa_compare_plt +
  ggtitle('HFA of each entry [kg/ha]') +
  theme(legend.position='none')

grid.arrange(grobs=list(p1, p2, p3),
             layout_matrix=rbind(c(1,1,1),
                                 c(2,2,2),
                                 c(NA,3,NA)))
```


```{r}
if (export) {
 out = "Figure S1 - Compare BLUP and average HFA.jpg" %>% 
    here('results', .)
  jpeg(out, width=8.5, height=5, units='in', res=300)
  grid.arrange(grobs=list(p1, p2, p3),
             layout_matrix=rbind(c(1,3),
                                 c(2,3)))
  dev.off()
}
```

